Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-06-15 submission/manuscript.Rmd
make discussion changes and remaining text changes suggested by Nick

**Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of taxa are responsible for the observed SCFA concentrations in patients with colorectal tumors.

@@

The observations from this study do not support the hypothesis that SCFA concentrations are lower in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3]. Further, no benefit could be found in using the concentrations to help classify individuals with and without tumors [Figure 2]. These observations are in stark contrast to some of the previous literature.

@@

One limitation of ours and others research into the effect of SCFAs in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Although there are limitations with the current research on SCFAs and colorectal tumors, evidence is building that these specific metabolites may not be protective. Yet, taxa that are associated with SCFA production are consistently higher in normal colons than those with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals with normal colons versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Protection could be via a different pathway and by extension other metabolites that have not been extensively studied. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results

@@

Our observations found no difference in SCFA concentration, their utility as a classification tool, and for genes for enzymes involved in SCFA synthesis between individuals with normal colons and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Overall, these results suggest that the SCFAs produced by resident microbes do not protect against tumor. By focusing on other types of metabolites, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

@@

**Figure 1. No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (A). Butyrate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (B). Propionate concentrations in fecal samples of individuals with normal colons, adenomas, and carcinomas (C). The black lines indicate the median SCFA concentration. Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D). Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F). The black dots and lines represent the median change in SCFA concentration.

>>>>

**Random Forest models with SCFA concentrations do not classify tumors better.** SCFA concentrations could improve prediction of tumors based on specific bacterial community structures. Our OTU data can be used in combination with SCFAs to assess whether there is a community dependent context to SCFA classification of tumors. Using the Random Forest algorithm we built models with OTU abundance data only or OTU abundances and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although including SCFA concentrations did not improve classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor. This would be consistent with previous 16S rRNA gene sequencing results where many taxa associated with SCFA production are decreased [@Shah2017; @meta_analysis_crc_Sze2018].

@@

The observations from this study do not support the hypothesis that SCFA concentrations are different in individuals with tumors. Whether we directly measured the SCFA concentration or investigated genes associated with their production, no difference could be identified [Figure 1 & 3]. There is an intriguing reason why taxa associated with SCFA production are decreased in CRC but the genes involoved with its' production are not. Mouth-associated microbes such as *Fusobacterium nucleatum* and *Porphyromonas asaccharolytica* have been found to be increased in patients with carcinomas versus individuals without tumors [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Both of these bacterial species are known to have strains that can produce SCFAs such as butyrate [@Vital2014]. Thus the reason we may be observing no change in genes involved with SCFA synthesis, as well as no change to SCFAs themselves, is because the production is being supported by more inflammatory microbes associated with CRC. Additionally, our observations that no benefit could be found in using the concentrations to help classify individuals with and without tumors would be consistent with this reason [Figure 2]. However, our observations are in stark contrast to some of the previous literature.

@@

One limitation of current research into the effect of SCFAs in CRC has been that all tumors are treated as the same type. However, there are known differences in the types of mutations that occur [@Fearon1990] and treating all tumors as equal may actually hide any benefit that could be found in specific subsets of individuals. Similar to the idea of specific immunotherapy for specific tumors [@Thomas2016], SCFAs may have beneficial effects for specific types of colorectal tumors. Future research will need to test if this is a valid hypothesis. Another limitation is that a fecal sample may not be an ideal type of biospecimen and that the effect SCFAs have on tumorigenesis is only detected in the colon. However, most *in vivo* studies as well as human studies have used fecal material in their analysis [@Ohigashi2013; @Bishehsari2018]. Additionally, studies that measure SCFA changes after fiber supplementation use fecal material to track these responses [@scfa_measures_venkataraman2016]. Although there are limitations with the current research on SCFAs and colorectal tumors, our observations along with the randomized controlled trials on fiber supplementation in tumor recurrence [@Yao2017] provide evidence that these specific metabolites may not be protective. Yet, taxa that are associated with SCFA production are consistently higher in indivdiuals without colon tumors than patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018].

The potential protection against colorectal cancers may not be from SCFAs even though taxa associated with their production are higher in individuals without tumors versus patients with carcinomas [@Shah2017; @normalization_Sze2017; @meta_analysis_crc_Sze2018]. Protection could be via a different pathway and by extension other metabolites that have not been extensively studied. Alternatively, protection may not occur via a metabolite but instead through niche exclusion of mouth-associated microbes (e.g. *Fusobacterium*, *Porphyromonas*, *Parvimonas*, *Peptostreptococcus* [@Zeller2014; @Baxter2016; @Kostic2011]). The idea of niche exclusion is similar to how the community protects against *Clostridium difficile* infection [@Theriot2015] with chronic inflammation replacing the role of antibiotics. Although we did not find lower concentrations of SCFAs associated with colorectal tumors, we think that there are many exciting new avenues to explore because of these results.

@@

Our observations found no difference in SCFA concentration, their utility as a classification tool, or for genes of enzymes involved in SCFA synthesis between individuals without colon tumors and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Overall, these results suggest that the SCFAs typically produced by resident microbes do not protect against tumor. By focusing on other types mechanisms, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

@@

**Figure 1. No change in SCFA measurements was observed between normal, adenoma, and carcinoma individuals using HPLC.** Acetate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (A). Butyrate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (B). Propionate concentrations in fecal samples of individuals without colon tumors, adenomas, and carcinomas (C). The black lines indicate the median SCFA concentration. Acetate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (D). Butyrate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (E). Propionate concentrations in fecal samples before and after treatment for adenoma (yellow) and carcinoma (red) (F). The black dots and lines represent the median change in SCFA concentration.
